---
title: "10x RNA-seq analysis of MVT1 cancer and microenvironment in tumor and mets"
subtitle: "Analysis of ligand-receptor interactions using LRdb from SingleCellSignalR and CellPhoneDB"
author: "Francesc Castro-Giner"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
  cellranger_count_dir: ./data/rnaseq/p26532_o28268/pipelines/cellranger_featurebarcodes
  lrdb_dir: ./output/p26532_o28268/p26532_o28268-10x_rnaseq-pll-lr-hq-lrdb
  cellphonedb_dir: ./output/p26532_o28268/p26532_o28268-10x_rnaseq-pll-lr-hq-cellphonedb
  ncores: 6
---


## Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	error = FALSE,
	fig.align = "center",
	fig.asp = 0.618,
	fig.showtext = FALSE,
	fig.width = 3.5,
	message = FALSE,
	warning = FALSE,
	dev = c("png", "pdf"),
	dpi = 600,
	engine.opts = list(bash = "-l"),
	results = "asis"
)
options(stringsAsFactors = FALSE)

use_seed <- 1100101
set.seed(use_seed)
```

Load packages
```{r load-libraries}
library(tidyverse)
library(showtext)
library(foreach)
library(DT)
library(knitr)
library(kableExtra)
library(cowplot)
library(colorblindr)
library(openxlsx)
library(ggpubr)
library(magrittr)

library(org.Mm.eg.db)
library(org.Hs.eg.db)
library(ComplexHeatmap)
library(circlize)
library(ggbeeswarm)
# library(arsenal)
library(RColorBrewer)
# library(ggpubr)
# library(pheatmap)
# library(patchwork)
# library(ggrepel)
# library(openxlsx)
# 
library(scater)
# library(DropletUtils)
# library(Seurat)
# library(scran)
# library(BiocSingular)
# library(batchelor)
# library(bluster)
# library(celldex)
# library(SingleR)
# library(speckle)
# library(miloR)
# library(clusterProfiler)
```

Set font family for figures
```{r set-font, eval = FALSE}
font_add("Helvetica", "./configuration/fonts/Helvetica.ttc")
showtext_auto()
```

Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
source("./configuration/rmarkdown/color_palettes.R")
```

Load custom functions
```{r load-functions}
iscore.permutation.test <- function(x, y, n){
  original <- avg_ligand*avg_receptor
  distribution <- foreach(i=1:n, .combine = c) %do% {
    x*(sample(y, length(y), FALSE))
  }
  foreach(o = original, .combine = c) %do% {
    sum(distribution >= o)/length(distribution)
  }
}
normalize01 <- function(x) {
  (x - min(x)) / ( max(x) -  min(x))
}
```

Clean files generated in previous runs
```{r clean-files}
rmd_file <- current_input()
if(!is.null(rmd_file)) {
  figures_dir <- file.path('./docs/figure',rmd_file)
  if(dir.exists(figures_dir)) {
    unlink(file.path(figures_dir, "*"))
  }
}

```

Other configuration
```{r conf-general}
# Site levels order
site_caps_ord <- c('Primary tumor', 'Liver', 'Lung')

#Select specific interactions
lrdb_selected_interacting_pair <- c(
  'Acvr1-Bmpr2-Bmp6',
  'Bmpr1a-Bmpr2-Bmp6',
  'Bmpr1a-Bmpr2-Bmp2',
  
  'Tnfrsf1a-Grn',
  'Tnfrsf1b-Grn',
  'Ide-Ccl9',  # Ccl9 and Ccl6 are othologous for CCL23
  'Ide-Ccl6',
  
  'Grn-Tnfrsf1a',
  'Grn-Tnfrsf1b',
  'Ccl9-Ide',  # Ccl9 and Ccl6 are othologous for CCL23
  'Ccl6-Ide',
  
  'Bmp2-Bmpr2',
  'Bmp2-Bmpr1a',
  'Bmp2-Acvr1',
  'Spp1-Itgb1'
  )

cellphonedb_selected_interacting_pair <- c(
  'Acvr1_Bmpr2_Bmp6',
  'Bmpr1a_Bmpr2_Bmp6',
  'Bmpr1a_Bmpr2_Bmp2',
  
  'Tnfrsf1a_Grn',
  'Tnfrsf1b_Grn',
  'Ide_Ccl9',  # Ccl9 and Ccl6 are othologous for CCL23
  'Ide_Ccl6',
  
  'Grn_Tnfrsf1a',
  'Grn_Tnfrsf1b',
  'Ccl9_Ide',  # Ccl9 and Ccl6 are othologous for CCL23
  'Ccl6_Ide'
  )

selected_interacting_pair <- c(
  lrdb_selected_interacting_pair,
  cellphonedb_selected_interacting_pair
) %>% unique
```

## Cytokine-receptor interaction
```{r lrdb-mouse-site-load}
iscore_df <- readRDS(file.path(params$lrdb_dir, 'cytokine_receptor_iscore.rds')) %>% 
  mutate(
    site = factor(site, site_caps_ord),
    source = 'LRdb'
    )
iscore_ll_ttest <- readRDS(file.path(params$lrdb_dir, 'cytokine_receptor_iscore_ll_ttest.rds'))
```


```{r cellphonedb-tme_merged-load}
cpd_merged <- readRDS(file = file.path(params$cellphonedb_dir, 
                                       'tme_merged-cpdb.rds')) %>% 
  mutate(
    site = factor(site, site_caps_ord),
    source = 'CPdb'
    )

cpd_ll_ttest <- readRDS(file = file.path(params$cellphonedb_dir, 
                                            'tme_merged-cpdb-ll-ttest.rds'))

# modify cpd to have same format as iscore_df
cpd_merged <- cpd_merged %>% 
   dplyr::rename(
     lr = interacting_pair,
     mouse_id = mouse, 
     iscore = mean,
     iscore_norm = mean_norm
   ) %>% 
   mutate(
     site_mouse = paste(site, mouse_id, sep = '_'),
     p.value = NA,
     p.value_scaled = NA,
     iscore_scaled = NA
   ) %>% 
   dplyr::select(all_of(colnames(iscore_df)))

# Combine with iscore data
iscore_df <- rbind(iscore_df, cpd_merged)
iscore_ll_ttest <- rbind(iscore_ll_ttest, cpd_ll_ttest)
```


### Stripchart - Selected interactions
```{r lrdb-cpdb-mouse-site-sel-stripchart, fig.dim=c(4.3078, 8.164469)}
use_data <- iscore_df %>% 
  mutate(
    mouse_id =  gsub('mouse_', '', mouse_id),
    site_mouse = paste(site, mouse_id, sep = '_')
  ) %>% 
  filter(lr %in% selected_interacting_pair)

use_lr <- intersect(selected_interacting_pair, use_data$lr)

# Manual P-values
lr_y_pos <- use_data %>% 
  filter(lr %in% use_lr) %>% 
  group_by(lr) %>% 
  summarise(y.max = max(iscore)) %>% 
  mutate(y.position = y.max + (0.10*y.max))

use_test <- iscore_ll_ttest %>% filter(lr %in% use_lr) %>% 
  dplyr::rename(group1=m_lung, group2=m_liver) %>% 
  mutate(
    group1 = 'Liver',
    group2 = 'Lung',
    # groups =  list(c('Liver', 'Lung')),
    # xmin = 1,
    # xmax = 2,
    p = format.pval(p, digits = 1)
  ) %>% 
  left_join(lr_y_pos)

  
use_data %>% 
  filter(lr %in% use_lr) %>%
  mutate(lr = factor(lr, levels = selected_interacting_pair)) %>% 
  ggplot(aes(site, iscore, size = -log10(p.value))) +
  geom_boxplot(alpha = 0.4, color = 'black', aes(fill = site), show.legend = FALSE, size = 0.25, fatten = 1) +
  geom_quasirandom(method="smiley", width = 0.1, alpha = 1, size = 2, shape = 21, color = 'black', aes(fill = site), stroke = 0.4) +
  # stat_compare_means(comparisons = list(c("Liver", "Lung")), method = 't.test') +
  stat_pvalue_manual(use_test,  label = "p = {p}", vjust = -0.5, label.size = geom_text_size) +
  # scale_fill_OkabeIto() +
  # scale_color_OkabeIto() +
  scale_fill_manual(values = site_palette) +
  scale_color_manual(values = site_palette) +
  scale_y_continuous(expand = c(0.05, 0, 0.35, 0)) +
  scale_x_discrete(expand = c(0, 0.5)) +
  facet_wrap(vars(lr), scales = 'free_y', ncol = 2) +
  theme(
    # axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank(),
    panel.border = element_rect(colour = "black", size=0.5),
    axis.line = element_blank(),
    strip.background = element_blank()
  ) +
  guides(fill = guide_legend(override.aes = list(size = 3))) +
  labs(
    fill = NULL, 
    x = NULL,
    y = 'Interaction score',
    caption = 'One-sided T-test P value'
  )

```



## Cytokine-receptor interaction by cell type

```{r lrdb-mouse-site-ct-load}
iscore_df <- readRDS(file.path(params$lrdb_dir, 'cytokine_receptor_celltype_iscore.rds')) %>% 
  mutate(
     site = factor(site, site_caps_ord),
     source = 'LRdb'
     )
iscore_ll_ttest <- readRDS(file.path(params$lrdb_dir, 'cytokine_receptor_celltype_iscore_ll_ttest.rds'))
```


```{r cpdb-mouse-site-ct-load}
cpd_merged <- readRDS(file = file.path(params$cellphonedb_dir, 'site_type-cpdb.rds')) %>% 
  mutate(
    site = factor(site, site_caps_ord),
    source = 'CPdb'
    ) %>% 
  filter(celltype_src == "immgen") %>% 
  filter(interacting_partner != "Cancer")

# modify cpd to have same format as iscore_df
cpd_merged <- cpd_merged %>% 
  dplyr::rename(
    lr = interacting_pair,
    cell_type = interacting_partner,
    mouse_id = mouse, 
    iscore = mean,
    iscore_norm = mean_norm
    ) %>% 
  mutate(
    site_mouse = paste(site, mouse_id, sep = '_'),
    ct_site_mouse =  paste(cell_type, site_mouse, sep = '_'),
    ct_lr = paste(cell_type, lr, sep = '_'),
    p.value = NA,
    p.value_scaled = NA
    ) %>% 
  dplyr::select(all_of(colnames(iscore_df)))
# Combine with iscore data
iscore_df <- rbind(iscore_df, cpd_merged)
```


### Dot plots all cell types
```{r lrdb-cpdb-mouse-site-all_ct-dotplot-conf}
use_data <- iscore_df %>% 
  group_by(lr, cell_type, site) %>% 
  summarise(miscore = median(iscore))

use_data_to_plot <- use_data %>%
  mutate(site = gsub("_", "\n", site)) %>%
  mutate(cell_type = gsub(" ", "\n", cell_type)) %>%
  filter(lr %in% selected_interacting_pair) %>% 
  mutate(lr = factor(lr, levels = selected_interacting_pair))
 
# Add empty levels
ipartner_levels <- unique(use_data_to_plot$cell_type)
site_levels <- unique(iscore_df$site)
 
new_data <- foreach(ipair = unique(use_data_to_plot$lr), .combine = rbind) %do% {
  foreach(ipartner = ipartner_levels, .combine = rbind) %do% {
    foreach(isite = site_levels, .combine = rbind) %do% {
      fdata <- use_data_to_plot %>% 
        filter(lr == ipair & cell_type == ipartner & isite == site)
      if(nrow(fdata) == 0){
        data.frame(lr = ipair, cell_type = ipartner, site = isite, miscore = 0)
        } else {
          fdata %>% data.frame
        }
    }
  }
}
 
use_data_to_plot <- new_data
```

#### Normal plot
```{r lrdb-cpdb-mouse-site-all_ct-dotplot, fig.dim=c(4.3078, 8.164469)}
use_data_to_plot %>% 
  ggplot(aes(cell_type, site, 
             size = miscore,
             color = miscore)) +
  geom_point() +
  scale_color_distiller(palette = "Purples", na.value = "grey70", direction = 1) +
  # scale_color_purpleslog10scaled +
  # scale_color_viridis_log10scaled +
  facet_wrap(lr~., ncol = 2) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
    axis.ticks.y=element_blank(),
    panel.border = element_rect(colour = "black", size=0.5),
    panel.grid.major.x = element_line(linewidth = 0.2, colour="grey60", linetype = 'dashed'),
    strip.background = element_blank()
  ) +
  labs(
    x = NULL,
    y = '',
    color = 'Median\nscore', 
    size = 'Median\nscore',
    caption = 'Showing median score across replicates'
  )
```

#### Color palette log10 scale

The breaks in the color palette are scaled using log10. In order to avoid -Inf (log10(0)) values, 0 were transformed into 0.1.
```{r lrdb-cpdb-mouse-site-all_ct-dotplot-log10palette, fig.dim=c(4.3078, 8.164469)}
scale_color_purpleslog10scaled <- scale_colour_gradientn(
  colors = scales::brewer_pal(palette = "Purples")(9),
  values = rev(1-log10(seq(1:10))),
  trans = 'log10'
)

use_data_to_plot %>% 
  ggplot(aes(cell_type, site, 
             size = miscore,
             color = miscore)) +
  geom_point() +
  scale_color_purpleslog10scaled +
  facet_wrap(lr~., ncol = 2) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
    axis.ticks.y=element_blank(),
    panel.border = element_rect(colour = "black", size=0.5),
    panel.grid.major.x = element_line(linewidth = 0.2, colour="grey60", linetype = 'dashed'),
    strip.background = element_blank()
  ) +
  labs(
    x = NULL,
    y = '',
    color = 'Median\nscore', 
    size = 'Median\nscore',
    caption = 'Showing median score across replicates'
  )
```


#### Color capped
In the plot below, the color scale is capped at 2, limiting all values above 2 into 2. The labels in the color legend need to be modifed, and 2 should become ≥ 2.
```{r lrdb-cpdb-mouse-site-all_ct-dotplot-color_capped, fig.dim=c(4.3078, 8.164469)}
use_data_to_plot %>% 
  mutate(miscore_color = ifelse(miscore >= 2, 2, miscore)) %>% 
  ggplot(aes(cell_type, site, 
             size = miscore,
             color = miscore_color)) +
  geom_point() +
  scale_color_distiller(palette = "Purples", na.value = "grey70", direction = 1) +
  # scale_color_purpleslog10scaled +
  # scale_color_viridis_log10scaled +
  facet_wrap(lr~., ncol = 2) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
    axis.ticks.y=element_blank(),
    panel.border = element_rect(colour = "black", size=0.5),
    panel.grid.major.x = element_line(linewidth = 0.2, colour="grey60", linetype = 'dashed'),
    strip.background = element_blank()
  ) +
  labs(
    x = NULL,
    y = '',
    color = 'Median\nscore', 
    size = 'Median\nscore',
    caption = 'Showing median score across replicates'
  )
```

