---
title: "USB and USZ metastatic BRCA patients"
subtitle: "Data analysis"
author: "Francesc Castro-Giner"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
params:
  date: '`r format(Sys.Date(), "%B %d, %Y")`'
---

## Load libraries, additional functions and data

Setup environment
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(results='asis', echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.align = 'center', fig.width = 3.5, fig.asp = 0.618, dpi = 600, dev = c("png", "pdf"))

options(stringsAsFactors = FALSE)

use_seed <- 1100101
set.seed(use_seed)
```

Load packages
```{r load-libraries}
library(tidyverse)
library(showtext)
library(foreach)
library(DT)
library(knitr)
library(kableExtra)
library(arsenal)
library(ComplexHeatmap)
library(viridis)
library(RColorBrewer)
library(circlize)
library(ggalluvial)
library(survival)
library(ranger)
library(ggfortify)
library(colorblindr)
library(ggpubr)
library(magrittr)
library(ggsurvfit)
library(gtsummary)
```

Set font family for figures
```{r set-font, eval = FALSE}
font_add("Helvetica", "./configuration/fonts/Helvetica.ttc")
showtext_auto()
```

Load ggplot theme
```{r ggplot-theme}
source("./configuration/rmarkdown/ggplot_theme.R")
```

Load palettes
```{r load-palettes}
source("./configuration/rmarkdown/color_palettes.R")
```

Clean files generated in previous runs
```{r clean-files}
rmd_file <- current_input()
if(!is.null(rmd_file)) {
  figures_dir <- file.path('./docs/figure',rmd_file)
  if(dir.exists(figures_dir)) {
    unlink(file.path(figures_dir, "*"))
  }
}

```

Load data
```{r load-data}
db_usb <- readRDS('./data/patients/usb_metastatic_brca_patients.rds')
db_usz <- readRDS('./data/patients/usz_metastatic_brca_patients.rds')
```

## Data wrangling
Integration of USB and USZ data
```{r db-integration}
# db_usb <- readRDS('./data/patients/usb_metastatic_brca_patients.rds')
# db_usz <- readRDS('./data/patients/usz_metastatic_brca_patients.rds')
# setdiff(colnames(db_usz), colnames(db_usb))

db_usb %<>% 
  # Rename variables
  dplyr::rename(metastasis_count_nb = metastasis_count) %>% 
  # Metastasis sites
  dplyr::rename(
    lymph_nodes = distant_limph_nodes_lymphangiosis,
    peritoneum = peritoneal_retro_peritoneal
  ) %>% 
  # Other variables
  mutate(
    time_from_event_to_loss_to_follow_up = NA,
    loss_to_follow_up = NA,
    surv_status = 'death',
    patient_number = paste0('usb_', patient_number),
    cohort = 'USB'
  ) %>% 
  # Time variables
  mutate_at(vars(starts_with('time_')), as.numeric)

db_usz %<>% 
  # Metastasis sites
  dplyr::mutate(
    age_at_primary_diagnosis = NA,
    other = ifelse(adrenal, TRUE, other),
    other = ifelse(adnexal, TRUE, other)
  ) %>% 
  dplyr::select(-adrenal, -adnexal) %>% 
  # Other variables
  mutate(
    patient_number = paste0('usz_', patient_number),
    cohort = 'USZ'
  ) %>% 
  # Time variables
  mutate_at(vars(starts_with('time_')), as.numeric)

use_cols <- colnames(db_usb)[colnames(db_usb) %in% colnames(db_usz)]

data <-rbind(db_usb[,use_cols], db_usz[,use_cols])
```

Add mono-metastasis to bone as first event
```{r dw-first-bone}
pat_met_bone_first <- data %>% 
  filter(metastasis_nb == 1 & bone & metastasis_count_nb == 1) %>% 
  pull(patient_number)

data %<>% 
  mutate(
    metastasis_bone_first = patient_number %in% pat_met_bone_first,
  )
```


Generate additional variables
```{r gen-add-var}
met_sites <- data %>% dplyr::select(lymph_nodes:other) %>% colnames()
```

Generate patient level data
```{r gen-patient-data}
patient_data <- data %>% dplyr::select(
  patient_number:her2,
  age_at_primary_diagnosis, 
  metastasis_lung_first:molecular_subtype, 
  loss_to_follow_up:cohort,
  time_from_primary_diagnosis_to_death,
  time_from_primary_diagnosis_to_first_met
  ) %>% unique

patient_data_sites <- data %>% 
  select(patient_number, lymph_nodes:other) %>%
  group_by(patient_number) %>% 
  summarise_at(
    vars(lymph_nodes:other), sum, na.rm = TRUE
  ) %>% 
  mutate_at(
    vars(lymph_nodes:other), function(x) x>0
  )

patient_data %<>% 
  left_join(patient_data_sites)
```

## Data tables

### Table : Patient characteristic
```{r db-table-patient}
patient_data %>% 
  select(patient_number, cohort,
         er, pr, hr, her2, molecular_subtype,
         age_at_primary_diagnosis,
         surv_status,
         time_from_primary_diagnosis_to_death,
         time_from_primary_diagnosis_to_first_met,
         metastasis_lung_first,
         metastasis_liver_first,
         lymph_nodes:other
         ) %>% 
  dplyr::rename(
    days_from_primary_diagnosis_to_death = time_from_primary_diagnosis_to_death,
    days_from_primary_diagnosis_to_first_met = time_from_primary_diagnosis_to_first_met
  ) %>%
  datatable(., 
            rownames = FALSE, 
            filter = 'top', 
            caption = 'Patients characteristics',
            extensions = 'Buttons', 
            options = list(
              dom = 'Blfrtip',
              buttons = c('csv', 'excel')
            )) %>% 
  formatRound(c('age_at_primary_diagnosis'), digits = 2) 
```


### Table : Metastasis events
In the table below each row is a clinical event (diagnosis, metastasis), so each patient is represented in multiple rows
```{r db-table-events}
data %>% 
  mutate(
    event = ifelse(
      is.na(time_from_previous_event),
      'inital_diagnosis',
      'metastasis'
    )
  ) %>% 
  select(
    patient_number, cohort,
    metastasis_nb,
    event,
    time_to_death,
    time_from_previous_event,
    next_event,
    time_to_next_event,
    lymph_nodes:other
  ) %>% 
  rename_with(., ~ gsub('time_', 'days_', .x)) %>% 
  datatable(., 
            rownames = FALSE, 
            filter = 'top', 
            caption = 'Metastasis events',
            extensions = 'Buttons', 
            options = list(
              dom = 'Blfrtip',
              buttons = c('csv', 'excel')
            ))
```

## Data exploration for integrity

```{r data-exploration-integrity}
tab_labels <- data %>% dplyr::select(-patient_number, -starts_with("metastasis_sites")) %>% colnames()
use_formula <- as.formula(paste0('cohort ~', paste(tab_labels, collapse = '+')))
res_tab <- tableby(use_formula, data=data) 
summary(res_tab, pfootnote = TRUE)
```


## Survival analysis Lung vs Liver (monometastasis)

Here we compare patients with first metastasis to Lung versus Liver. Only monometastasis are accepted.

### Time to next event

Here we evaluate the time to the next event (metastasis or death)
```{r met-lung-liver-mono-time2next, fig.asp = 0.9, fig.width=5.2}
use_data <- data %>% 
  filter(metastasis_nb == 1) %>% 
  mutate(
    date_of_study_end = ifelse(cohort == 'USZ', "2023-03-01", NA),
    date_of_study_end = as.Date(date_of_study_end),
    time_to_study_end=  date_of_study_end - date_of_metastasis,
    time = time_to_next_event,
    time = ifelse(is.na(time), time_from_event_to_loss_to_follow_up, time),
    status = ifelse(next_event == 'cut_off' | next_event == 'loss_to_follow_up', 0, 1)
  )

use_data_lung <- use_data %>%  filter(metastasis_lung_first) %>% mutate(caterogy = 'Lung')
use_data_liver <- use_data %>%  filter(metastasis_liver_first) %>% mutate(caterogy = 'Liver')
use_data <- rbind(use_data_lung, use_data_liver) %>% 
  mutate(
    cohort_caterogy = paste0(cohort, '-', caterogy),
    caterogy = factor(caterogy, levels = c('Lung', 'Liver'))
    )
  

sfit <- survfit(Surv(time, status) ~ caterogy, data=use_data)
sdiff <- survdiff(Surv(time, status) ~ caterogy, data=use_data)
sfit2 <- survfit2(Surv(time, status) ~ caterogy, data=use_data)
coxph_res <- coxph(Surv(time, status) ~ caterogy, data=use_data)
# broom::tidy(sdiff)
# broom::glance(sdiff)

ggsurvfit(sfit2) +
  labs(
    x = "Days from metastasis diagnosis",
    y = "Overall survival probability",
    caption = paste("log-rank p-value =", format.pval(sdiff$pvalue, digits = 2))
  ) + 
  # add_confidence_interval() +
  add_risktable() +
  scale_color_manual(values = site_palette) +
  theme_project

summary(sfit)$table %>%
  kbl(caption = 'Kaplan-Meier estimates.') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)



coxph_res %>% broom::tidy() %>% 
  mutate(
    ci.l = estimate - 1.96*std.error,
    ci.u = estimate + 1.96*std.error 
  ) %>% 
  dplyr::rename(HR = estimate) %>% 
  mutate(
    HR = exp(HR),
    ci.u = exp(ci.u),
    ci.l = exp(ci.l),
    term = gsub('caterogy', '', term),
    p.value = format.pval(p.value)) %>% 
  kable(caption = "Cox PH model") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```


### Time to death
```{r met-lung-liver-mono-time2death, fig.asp = 0.9, fig.width=5.2}
use_data <- data %>% 
  filter(metastasis_nb == 1) %>% 
  mutate(
    date_of_study_end = ifelse(cohort == 'USZ', "2023-03-01", NA),
    date_of_study_end = as.Date(date_of_study_end),
    time_to_study_end=  date_of_study_end - date_of_metastasis,
    time = time_to_death,
    time = ifelse(surv_status == 'alive', time_to_study_end, time),
    time = ifelse(is.na(time), time_from_event_to_loss_to_follow_up, time),
    status = ifelse(is.na(time_to_death), 0, 1)
  )

use_data_lung <- use_data %>%  filter(metastasis_lung_first) %>% mutate(caterogy = 'Lung')
use_data_liver <- use_data %>%  filter(metastasis_liver_first) %>% mutate(caterogy = 'Liver')
use_data <- rbind(use_data_lung, use_data_liver) %>% 
  mutate(
    cohort_caterogy = paste0(cohort, '-', caterogy),
    caterogy = factor(caterogy, levels = c('Lung', 'Liver'))
    )

sfit <- survfit(Surv(time, status) ~ caterogy, data=use_data)
sdiff <- survdiff(Surv(time, status) ~ caterogy, data=use_data)
sfit2 <- survfit2(Surv(time, status) ~ caterogy, data=use_data)
coxph_res <- coxph(Surv(time, status) ~ caterogy, data=use_data)

# broom::tidy(sdiff)
# broom::glance(sdiff)

ggsurvfit(sfit2) +
  labs(
    x = "Days from metastasis diagnosis",
    y = "Overall survival probability",
    caption = paste("log-rank p-value =", format.pval(sdiff$pvalue, digits = 2))
  ) + 
  # add_confidence_interval() +
  add_risktable() +
  scale_color_manual(values = site_palette) +
  theme_project


summary(sfit2)$table %>%
  kbl(caption = 'Kaplan-Meier estimates.') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)


coxph_res %>% broom::tidy() %>% 
  mutate(
    ci.l = estimate - 1.96*std.error,
    ci.u = estimate + 1.96*std.error 
  ) %>% 
  dplyr::rename(HR = estimate) %>% 
  mutate(
    HR = exp(HR),
    ci.u = exp(ci.u),
    ci.l = exp(ci.l),
    term = gsub('caterogy', '', term),
    p.value = format.pval(p.value)) %>% 
  kable(caption = "Cox PH model") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```


## Survival analysis Bone vs Liver (monometastasis)

Here we compare patients with first metastasis to Lung versus Liver. Only monometastasis are accepted.

### Time to next event

Here we evaluate the time to the next event (metastasis or death)
```{r met-bone-liver-mono-time2next, fig.asp = 0.9, fig.width=5.2}
use_data <- data %>% 
  filter(metastasis_nb == 1) %>% 
  mutate(
    date_of_study_end = ifelse(cohort == 'USZ', "2023-03-01", NA),
    date_of_study_end = as.Date(date_of_study_end),
    time_to_study_end=  date_of_study_end - date_of_metastasis,
    time = time_to_next_event,
    time = ifelse(is.na(time), time_from_event_to_loss_to_follow_up, time),
    status = ifelse(next_event == 'cut_off' | next_event == 'loss_to_follow_up', 0, 1)
  )

use_data_bone <- use_data %>%  filter(metastasis_bone_first) %>% mutate(caterogy = 'Bone')
use_data_liver <- use_data %>%  filter(metastasis_liver_first) %>% mutate(caterogy = 'Liver')
use_data <- rbind(use_data_bone, use_data_liver) %>% 
  mutate(
    cohort_caterogy = paste0(cohort, '-', caterogy),
    caterogy = factor(caterogy, levels = c('Liver', 'Bone'))
    )
  

sfit <- survfit(Surv(time, status) ~ caterogy, data=use_data)
sdiff <- survdiff(Surv(time, status) ~ caterogy, data=use_data)
sfit2 <- survfit2(Surv(time, status) ~ caterogy, data=use_data)
coxph_res <- coxph(Surv(time, status) ~ caterogy, data=use_data)
# broom::tidy(sdiff)
# broom::glance(sdiff)

ggsurvfit(sfit2) +
  labs(
    x = "Days from metastasis diagnosis",
    y = "Overall survival probability",
    caption = paste("log-rank p-value =", format.pval(sdiff$pvalue, digits = 2))
  ) + 
  # add_confidence_interval() +
  add_risktable() +
  scale_color_manual(values = site_palette) +
  theme_project

summary(sfit)$table %>%
  kbl(caption = 'Kaplan-Meier estimates.') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)



coxph_res %>% broom::tidy() %>% 
  mutate(
    ci.l = estimate - 1.96*std.error,
    ci.u = estimate + 1.96*std.error 
  ) %>% 
  dplyr::rename(HR = estimate) %>% 
  mutate(
    HR = exp(HR),
    ci.u = exp(ci.u),
    ci.l = exp(ci.l),
    term = gsub('caterogy', '', term),
    p.value = format.pval(p.value)) %>% 
  kable(caption = "Cox PH model") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```


### Time to death: 5-years interval
```{r met-bone-liver-mono-time2death-5y, fig.asp = 0.9, fig.width=5.2}
interval_thrs_days <- 5*365
use_data <- data %>% 
  filter(metastasis_nb == 1) %>% 
  mutate(
    date_of_study_end = ifelse(cohort == 'USZ', "2023-03-01", NA),
    date_of_study_end = as.Date(date_of_study_end),
    time_to_study_end=  date_of_study_end - date_of_metastasis,
    time = time_to_death,
    time = ifelse(surv_status == 'alive', time_to_study_end, time),
    time = ifelse(is.na(time), time_from_event_to_loss_to_follow_up, time),
    status = ifelse(is.na(time_to_death), 0, 1),
    
    interval_censor = time > interval_thrs_days,
    time = ifelse(interval_censor, interval_thrs_days, time),
    status = ifelse(interval_censor, 0, status)
  )

use_data_bone <- use_data %>%  filter(metastasis_bone_first) %>% mutate(caterogy = 'Bone')
use_data_liver <- use_data %>%  filter(metastasis_liver_first) %>% mutate(caterogy = 'Liver')
use_data <- rbind(use_data_bone, use_data_liver)  %>% 
  mutate(
    cohort_caterogy = paste0(cohort, '-', caterogy),
    caterogy = factor(caterogy, levels = c('Liver', 'Bone'))
    )

sfit <- survfit(Surv(time, status) ~ caterogy, data=use_data)
sdiff <- survdiff(Surv(time, status) ~ caterogy, data=use_data)
sfit2 <- survfit2(Surv(time, status) ~ caterogy, data=use_data)
coxph_res <- coxph(Surv(time, status) ~ caterogy, data=use_data)

# broom::tidy(sdiff)
# broom::glance(sdiff)

ggsurvfit(sfit2) +
  labs(
    x = "Days from metastasis diagnosis",
    y = "Overall survival probability",
    caption = paste("log-rank p-value =", format.pval(sdiff$pvalue, digits = 2))
  ) + 
  # add_confidence_interval() +
  add_risktable() +
  scale_color_manual(values = site_palette) +
  theme_project

summary(sfit2)$table %>%
  kbl(caption = 'Kaplan-Meier estimates.') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)


coxph_res %>% broom::tidy() %>% 
  mutate(
    ci.l = estimate - 1.96*std.error,
    ci.u = estimate + 1.96*std.error 
  ) %>% 
  dplyr::rename(HR = estimate) %>% 
  mutate(
    HR = exp(HR),
    ci.u = exp(ci.u),
    ci.l = exp(ci.l),
    term = gsub('caterogy', '', term),
    p.value = format.pval(p.value)) %>% 
  kable(caption = "Cox PH model") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)

```


## Survival analysis Bone vs Lung (monometastasis)

Here we compare patients with first metastasis to Lung versus Lung. Only monometastasis are accepted.

### Time to next event

Here we evaluate the time to the next event (metastasis or death)
```{r met-bone-lung-mono-time2next, fig.asp = 0.9, fig.width=5.2}
use_data <- data %>% 
  filter(metastasis_nb == 1) %>% 
  mutate(
    date_of_study_end = ifelse(cohort == 'USZ', "2023-03-01", NA),
    date_of_study_end = as.Date(date_of_study_end),
    time_to_study_end=  date_of_study_end - date_of_metastasis,
    time = time_to_next_event,
    time = ifelse(is.na(time), time_from_event_to_loss_to_follow_up, time),
    status = ifelse(next_event == 'cut_off' | next_event == 'loss_to_follow_up', 0, 1)
  )

use_data_bone <- use_data %>%  filter(metastasis_bone_first) %>% mutate(caterogy = 'Bone')
use_data_lung <- use_data %>%  filter(metastasis_lung_first) %>% mutate(caterogy = 'Lung')
use_data <- rbind(use_data_bone, use_data_lung) %>% 
  mutate(
    cohort_caterogy = paste0(cohort, '-', caterogy),
    caterogy = factor(caterogy, levels = c('Lung', 'Bone'))
    )
  

sfit <- survfit(Surv(time, status) ~ caterogy, data=use_data)
sdiff <- survdiff(Surv(time, status) ~ caterogy, data=use_data)
sfit2 <- survfit2(Surv(time, status) ~ caterogy, data=use_data)
coxph_res <- coxph(Surv(time, status) ~ caterogy, data=use_data)
# broom::tidy(sdiff)
# broom::glance(sdiff)

ggsurvfit(sfit2) +
  labs(
    x = "Days from metastasis diagnosis",
    y = "Overall survival probability",
    caption = paste("log-rank p-value =", format.pval(sdiff$pvalue, digits = 2))
  ) + 
  # add_confidence_interval() +
  add_risktable() +
  scale_color_manual(values = site_palette) +
  theme_project

summary(sfit)$table %>%
  kbl(caption = 'Kaplan-Meier estimates.') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)



coxph_res %>% broom::tidy() %>% 
  mutate(
    ci.l = estimate - 1.96*std.error,
    ci.u = estimate + 1.96*std.error 
  ) %>% 
  dplyr::rename(HR = estimate) %>% 
  mutate(
    HR = exp(HR),
    ci.u = exp(ci.u),
    ci.l = exp(ci.l),
    term = gsub('caterogy', '', term),
    p.value = format.pval(p.value)) %>% 
  kable(caption = "Cox PH model") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```


### Time to death: 5-years interval
```{r met-bone-lung-mono-time2death-5y, fig.asp = 0.9, fig.width=5.2}
interval_thrs_days <- 5*365
use_data <- data %>% 
  filter(metastasis_nb == 1) %>% 
  mutate(
    date_of_study_end = ifelse(cohort == 'USZ', "2023-03-01", NA),
    date_of_study_end = as.Date(date_of_study_end),
    time_to_study_end=  date_of_study_end - date_of_metastasis,
    time = time_to_death,
    time = ifelse(surv_status == 'alive', time_to_study_end, time),
    time = ifelse(is.na(time), time_from_event_to_loss_to_follow_up, time),
    status = ifelse(is.na(time_to_death), 0, 1),
    
    interval_censor = time > interval_thrs_days,
    time = ifelse(interval_censor, interval_thrs_days, time),
    status = ifelse(interval_censor, 0, status)
  )

use_data_bone <- use_data %>%  filter(metastasis_bone_first) %>% mutate(caterogy = 'Bone')
use_data_lung <- use_data %>%  filter(metastasis_lung_first) %>% mutate(caterogy = 'Lung')
use_data <- rbind(use_data_bone, use_data_lung)  %>% 
  mutate(
    cohort_caterogy = paste0(cohort, '-', caterogy),
    caterogy = factor(caterogy, levels = c('Lung', 'Bone'))
    )

sfit <- survfit(Surv(time, status) ~ caterogy, data=use_data)
sdiff <- survdiff(Surv(time, status) ~ caterogy, data=use_data)
sfit2 <- survfit2(Surv(time, status) ~ caterogy, data=use_data)
coxph_res <- coxph(Surv(time, status) ~ caterogy, data=use_data)

# broom::tidy(sdiff)
# broom::glance(sdiff)

ggsurvfit(sfit2) +
  labs(
    x = "Days from metastasis diagnosis",
    y = "Overall survival probability",
    caption = paste("log-rank p-value =", format.pval(sdiff$pvalue, digits = 2))
  ) + 
  # add_confidence_interval() +
  add_risktable() +
  scale_color_manual(values = site_palette) +
  theme_project

summary(sfit2)$table %>%
  kbl(caption = 'Kaplan-Meier estimates.') %>%
  kable_paper(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)


coxph_res %>% broom::tidy() %>% 
  mutate(
    ci.l = estimate - 1.96*std.error,
    ci.u = estimate + 1.96*std.error 
  ) %>% 
  dplyr::rename(HR = estimate) %>% 
  mutate(
    HR = exp(HR),
    ci.u = exp(ci.u),
    ci.l = exp(ci.l),
    term = gsub('caterogy', '', term),
    p.value = format.pval(p.value)) %>% 
  kable(caption = "Cox PH model") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)

```


